Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-27 astronum_2018/astronum-xrb.tex
some algorithm work

The 2D calculations shown in the previous section require about 10k
node-hours of computational time to get to about 10~ms (those
particular simulations were run at NERSC).  The GPU version of the
Castro hydrodynamics solver, under active development, is about
$10\times$ faster on a comparable node, so a GPU calculation could
bring the 2D cost down to 1k node-hours.  Most of this time is spent
in the reactions\MarginPar{verify fraction}.  A 3D version of this
calculation is estimated to require 5M node-hours on CPUs or 0.5M
node-hours using GPUs---this is about the size of an
annual \MarginPar{double check \#s} allocation.  Additionally, this is
still with the boosted flame, but we expect that once the flame is
developed, the boosting will not be as critical, due to hydrodynamic
effects.  We plan to test this soon.  In any case, this is less than
1\% of the neutron star surface, so a fully resolved calculation of
burning over the entire neutron star is out of reach, and will be for
some time.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  We can expect GPUs to perform 10--20$\times$ faster than
CPUs on modern supercomputer nodes.  This performance increase will
directly benefit XRB simulations.  GPU offloading of the reaction networks
will also allow us to explore larger networks and more detailed nuclear
physics.

If we continue to resolve the flame structure, then we need algroithms
that can do so less expensively.  We can be smarter about our grid
refinement strategy, putting the finest grids only at the very base of
the burning layer, where the flame is the thinnest.  Using
well-balanced schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium (HSE)
in the lower refined regions, reducing the number of zones needed to
maintain fidelity.  We can also explore initializing newly created fine
grids ahead of the burning with an understanding of hydrostatic
equilibrium to allow us to only refine the peak burning as the flame
moves across the star.

>>>>

The 2D calculations shown in the previous section require about 10k
node-hours of computational time to get to about 10~ms (those
particular simulations were run at NERSC).  Most of this time is spent
in the reactions\MarginPar{verify fraction}.  We really need to do 3D.
There the shear from the hurricane structure acting along the flame
front may cause instabilities that can affect the flame propagation.
This is something we can test with our current simulation methodology.
A 3D version of this calculation is estimated to require 5M node-hours
on CPUs---this is about the size of an annual \MarginPar{double check
  \#s} allocation.  Additionally, this is still with the boosted
flame, but we expect that once the flame is developed, the boosting
will not be as critical, due to hydrodynamic effects.  We plan to test
this soon.  We can bootstrap the 3D calculation by mapping the 2D
axisymmetric calculation, once the flame is firmly established, into
3D, rotating about the symmetry axis.  This will cut down the
computational time by about one-half.  In any case, this is less than
1\% of the neutron star surface, so a fully resolved calculation of
burning over the entire neutron star is out of reach, and will be for
some time.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores.  We see similar speed ups with reaction networks, with
these GPU ports under active development.  This means a GPU
calculation could lower the above costs by a factor of 10.  GPU
offloading of the reaction networks will also allow us to explore
larger networks and more detailed nuclear physics.

If we continue to resolve the flame structure, then we need algorithms
that can do so less expensively.  We can be smarter about our grid
refinement strategy, putting the finest grids only at the very base of
the burning layer, where the flame is the thinnest.  Using
well-balanced schemes~\cite{kappeli:2016} can help us retain
hydrostatic equilibrium (HSE) in the lower refined regions, reducing
the number of zones needed to maintain fidelity.  We can also explore
initializing newly created fine grids ahead of the burning with an
understanding of hydrostatic equilibrium to allow us to only refine
the peak burning as the flame moves across the star.
